---
title: "Big Data Bowl"
author: "Alex Veroulis"
date: "10/26/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading in data and making small edits
```{r}

setwd("~/Desktop/nfl-big-data-bowl-2022")

library(dplyr)

# player data
players <- read.csv("players.csv")

# game data
games <- read.csv("games.csv")

# DATA1: play data; 19979 plays total to start
plays <- read.csv("plays.csv") 

full_plays <- left_join(plays,games,by="gameId")

# DATA2: only include punts
plays_punt <- full_plays %>% dplyr::filter(specialTeamsPlayType == "Punt")

# looking at punt results
summary(as.factor(plays_punt$specialTeamsResult))

# look for kinds of penalties; notice there are 5350 plays without penalties
summary(as.factor(plays_punt$penaltyCodes))

# there are 641 plays with penalties; penalty (641) and non-penalty (5350) plays add up to total number of plays (5991)
length(which(as.factor(plays_punt$penaltyCodes) != "NA"))

# storing indices for penalty plays
penalty_index <- which(as.factor(plays_punt$penaltyCodes) != "NA")

# DATA3: non-penalty punts
punt_clean <- plays_punt[-penalty_index,]
summary(as.factor(punt_clean$specialTeamsResult))


# DATA4: filtering by desired results; total of 4330 plays kept
punt_kicks <- punt_clean %>% filter(specialTeamsResult %in% c("Downed","Fair Catch","Muffed","Return"))
  
summary(as.factor(punt_kicks$specialTeamsResult))

# making the response variable binary
punt_kicks$Response[punt_kicks$specialTeamsResult %in% c("Downed","Fair Catch")] <- "No"
punt_kicks$Response[punt_kicks$specialTeamsResult %in% c("Muffed","Return")] <- "Yes"

# PFF data
pff <- read.csv("PFFScoutingData.csv")

#join new punts dataset with pff data by gameId and playId
punt_pff <- left_join(punt_kicks,pff,by=c("gameId","playId"))


# 2018 tracking data
trk18 <- read.csv("tracking2018.csv")


# 2019 tracking data
trk19 <- read.csv("tracking2019.csv")


# 2020 tracking data
trk20 <- read.csv("tracking2020.csv")

# play catalogs for each individual season with tracking data
join2018 <- inner_join(punt_pff,trk18,by=c("gameId","playId"))
join2019 <- inner_join(punt_pff,trk19,by=c("gameId","playId"))
join2020 <- inner_join(punt_pff,trk20,by=c("gameId","playId"))

# plays catalogs for all season from 2018-2020
join_punts <- rbind(join2018,join2019)
join_punts2 <- rbind(join_punts,join2020)

```

# Reducing data
```{r}
summary(join_punts2)
library(dplyr)
library(tidyr)

# taking out unnecessary variables
join_punts3 <- join_punts2 %>% dplyr::select(-c(kickBlockerId,penaltyCodes,penaltyJerseyNumbers,penaltyYards,passResult,kickoffReturnFormation))

# decided to take out muffed punts; leaves us with 4196 plays
join_punts4 <- join_punts3 %>% dplyr::filter(specialTeamsResult != "Muffed")
summary(as.factor(join_punts4$specialTeamsResult))
summary(as.factor(join_punts4$event))
summary(as.factor(join_punts4$Response))

```

# Inspection of play events
```{r}
# Checking for miscellaneous events that don't belong

# checking for event == kickoff; disregard as none
u1 <- join_punts4 %>% dplyr::filter(event == "kickoff") %>% dplyr::select(c(gameId,playId)) %>% distinct()
up1 <- join_punts4 %>% dplyr::filter(gameId=="2019100600",playId=="2337")
summary(as.factor(up1$event))

# checking for event == autoevent_kickoff; disregard as none
u2 <- join_punts4 %>% dplyr::filter(event == "autoevent_kickoff") %>% dplyr::select(c(gameId,playId)) %>% distinct()

up2 <- join_punts4 %>% dplyr::filter(gameId=="2020091304",playId=="116")
summary(as.factor(up2$event))

# checking for event == kick_received; disregard as none
u3 <- join_punts4 %>% dplyr::filter(event == "kick_received") %>% dplyr::select(c(gameId,playId)) %>% distinct()

up3 <- join_punts4 %>% dplyr::filter(gameId=="2018090903",playId=="1844") 

# checking for event == kickoff_land (which is considered the same as punt_land, change to punt_land)
up4 <- join_punts4 %>% dplyr::filter(gameId=="2018123015",playId=="2651")

# checking for event == out_of_bounds; this is ok
up5 <- join_punts4 %>% dplyr::filter(gameId=="2018090901",playId=="2632")

# checking for event == pass_forward; change to none
u6 <- join_punts4 %>% dplyr::filter(event == "pass_forward") %>% dplyr::select(c(gameId,playId)) %>% distinct()
up6 <- join_punts4 %>% dplyr::filter(gameId == "2020111512",playId=="2109")

# checking for event == play_action; change to none
u7 <- join_punts4 %>% dplyr::filter(event == "play_action") %>% dplyr::select(c(gameId,playId)) %>% distinct()
up7 <- join_punts4 %>% dplyr::filter(gameId == "2018093011",playId=="168")

# checking for event == punt_blocked
punt_blocks <- join_punts4 %>% dplyr::filter(event=="punt_blocked")
# getting game and play Ids for these plays
punt_block_ids <- punt_blocks %>% dplyr::select(c(gameId,playId)) %>% distinct()
# extracting all frames for the specified plays
pbsample <- join_punts4 %>% dplyr::filter(gameId %in% punt_block_ids$gameId & playId %in% punt_block_ids$playId)

# REMOVING PUNT BLOCKS FROM join_punts4 
join_punts5 <- join_punts4 %>% dplyr::filter(!(playDescription %in% pbsample$playDescription))
# results in 4187 total plays
jp5 <- join_punts5 %>% dplyr::select(gameId,playId) %>% distinct()
summary(as.factor(join_punts5$event))


# checking for event == punt_muffed
punt_muffs <- join_punts5 %>% dplyr::filter(event=="punt_muffed")
# getting game and play Ids for these plays
punt_muffs_ids <- punt_muffs %>% dplyr::select(c(gameId,playId)) %>% distinct()
# extracting all frames for the specified plays
pmsample <- join_punts5 %>% dplyr::filter(gameId %in% punt_muffs_ids$gameId & playId %in% punt_muffs_ids$playId)

# REMOVING MUFFS FROM join_punts5 to get final dataset
join_punts6 <- join_punts5 %>% dplyr::filter(!(playDescription %in% pmsample$playDescription))
# results in 4185 plays
jp6 <- join_punts6 %>% dplyr::select(gameId,playId) %>% distinct()

# checking for event == safety; this is ok since it's after the catch
up8 <- join_punts6 %>% dplyr::filter(gameId=="2020122012",playId=="2492")

summary(as.factor(join_punts6$event))

#changing kickoff_land to punt_land
join_punts6$event <- as.factor(join_punts6$event)
levels(join_punts6$event)[levels(join_punts6$event)=="kickoff_land"] <- "punt_land"

# changing events to NONE
levels(join_punts6$event)[levels(join_punts6$event)=="kickoff"] <- "None"
levels(join_punts6$event)[levels(join_punts6$event)=="autoevent_kickoff"] <- "None"
levels(join_punts6$event)[levels(join_punts6$event)=="kick_received"] <- "None"
levels(join_punts6$event)[levels(join_punts6$event)=="pass_forward"] <- "None"
levels(join_punts6$event)[levels(join_punts6$event)=="play_action"] <- "None"


```


# Looking at the distribution of important frames and editing plays that have mislabeled events
```{r}
# critical frames

# snaps (4156 plays)
join_ballsnap <- join_punts6 %>% dplyr::filter(event == "ball_snap")
plays_ballsnap <- join_ballsnap %>% dplyr::select(gameId,playId) %>% distinct()

#punts (4180 plays)
join_puntplays <- join_punts6 %>% dplyr::filter(event == "punt")
plays_puntplays <- join_puntplays %>% dplyr::select(gameId,playId) %>% distinct()

#fair catches (1534 plays)
join_fair <- join_punts6 %>% dplyr::filter(event == "fair_catch")
plays_fair <- join_fair %>% dplyr::select(gameId,playId) %>% distinct()


#punt received (1899 plays)
join_received <- join_punts6 %>% dplyr::filter(event == "punt_received")
plays_received <- join_received %>% dplyr::select(gameId,playId) %>% distinct()

#punt land (838 plays)
join_land <- join_punts6 %>% dplyr::filter(event == "punt_land")
plays_land <- join_land %>% dplyr::select(gameId,playId) %>% distinct()


## INTERSECTIONS

# 1.) fair catch and punt land intersection; punt_land as an erroneous event, must change to None
fc_land <- inner_join(plays_land,plays_fair) # 2 plays

join_punts6[which(join_punts6$gameId %in% fc_land$gameId & join_punts6$playId %in% fc_land$playId & join_punts6$event == "punt_land"),]$event <- "None"


#############

# 2.) fair catch and punt received intersection; punt_received as an erroneous event, must change to None
fc_rec <- inner_join(plays_fair,plays_received) # 14 plays

join_punts6[which(join_punts6$gameId %in% fc_rec$gameId & join_punts6$playId %in% fc_rec$playId & join_punts6$event == "punt_received"),]$event <- "None"


#############

# 3.) punt received and punt land intersection; looking for plays with a bounced return where a punt is received after the ball lands (127 plays), reassign as "no" for response variable
received_land <- inner_join(plays_received,plays_land)

# changing response variable to "No" for bounced returns, only considering the initial decision to catch/return or not 
join_punts6[which(join_punts6$gameId %in% received_land$gameId & join_punts6$playId %in% received_land$playId),]$Response <- "No"

# also have to change "punt received" events to "None" since we are only focusing on the punt landing
join_punts6[which(join_punts6$gameId %in% received_land$gameId & join_punts6$playId %in% received_land$playId & join_punts6$event == "punt_received"),]$event <- "None"


```


# Creating and editing necessary variables
```{r}
# separating gunners
final_ds <- join_punts6 %>% separate(gunners, c("gun1","gun2"),sep="; ") 

# adding new variables to identify coverage and kick team members
final_ds2 <- final_ds %>% mutate(tmabr =
case_when(
  team == "home" ~ homeTeamAbbr,
  team == "away" ~ visitorTeamAbbr,
  team == "football" ~ "football"
),
kr = ifelse(possessionTeam == tmabr,'kick','return'),
tm_num = paste(tmabr,jerseyNumber),
gunner_status = 
  case_when(
    tm_num == gun1 ~ 1,
    tm_num == gun2 ~ 1,
    TRUE ~ 0
  ),returner_status = ifelse(returnerId == nflId,1,0))

# changing classes of variables to factors
final_ds2$quarter <- as.factor(final_ds2$quarter)
final_ds2$season <- as.factor(final_ds2$season)
final_ds2$Response <- as.factor(final_ds2$Response)
final_ds2$snapDetail <- as.factor(final_ds2$snapDetail)
final_ds2$kickType <- as.factor(final_ds2$kickType)
final_ds2$kickDirectionIntended <- as.factor(final_ds2$kickDirectionIntended)
final_ds2$kickDirectionActual <- as.factor(final_ds2$kickDirectionActual)
final_ds2$kickContactType <- as.factor(final_ds2$kickContactType)
final_ds2$position <- as.factor(final_ds2$position)
final_ds2$team <- as.factor(final_ds2$team)
final_ds2$playDirection <- as.factor(final_ds2$playDirection)
final_ds2$kr <- as.factor(final_ds2$kr)
final_ds2$gunner_status <- as.factor(final_ds2$gunner_status)
final_ds2$returner_status <- as.factor(final_ds2$returner_status)


# summary of response variable
final_ds2plays <- final_ds2 %>% dplyr::select(gameId,playId,Response) %>% distinct()
summary(final_ds2plays$Response) # 1746 returns, 2439 non-returns

#change gameClock variable
library(lubridate)
library(stringr)
final_ds2$gameClock <- substr(final_ds2$gameClock,1,5)
final_ds2$gameClock <- lubridate::ms(final_ds2$gameClock)
gcindex <- which(final_ds2$quarter %in% c(1,3))
final_ds2$gameClock[gcindex] <- final_ds2$gameClock[gcindex] + lubridate::ms("15M 0S")
summary(final_ds2$gameClock)

# creating kickDirectionMatch variable (did the kick go in the intended direction?)
final_ds2$kickDirectionMatch <- ifelse(final_ds2$kickDirectionIntended == final_ds2$kickDirectionActual,"yes","no")
final_ds2$kickDirectionMatch <- as.factor(final_ds2$kickDirectionMatch)

# creating scorediff variable (from perspective of returning team)
final_ds2$scorediff <- numeric(nrow(final_ds2))

# counting home team score as the return team score
hrindex <- which(final_ds2$team == "home" & final_ds2$kr == "return")
akindex <- which(final_ds2$team == "away" & final_ds2$kr == "kick")
# score differential for home returners
final_ds2$scorediff[c(hrindex,akindex)] <- final_ds2$preSnapHomeScore[c(hrindex,akindex)] - final_ds2$preSnapVisitorScore[c(hrindex,akindex)]

# counting away team score as the return team score
hkindex <- which(final_ds2$team == "home" & final_ds2$kr == "kick")
arindex <- which(final_ds2$team == "away" & final_ds2$kr == "return")
# score differential for away returners
final_ds2$scorediff[c(hkindex,arindex)] <- final_ds2$preSnapVisitorScore[c(hkindex,arindex)] - final_ds2$preSnapHomeScore[c(hkindex,arindex)]

# team == football as having NA score difference
fnindex <- which(final_ds2$team == "football")
# score differential as NA for tracking the football
final_ds2$scorediff[fnindex] <- NA

# converting scorediff to a numeric variable
final_ds2$scorediff <- as.numeric(final_ds2$scorediff)

# for simplicity, taking out frames capturing the football's movement
final_ds2 <- final_ds2 %>% dplyr::filter(team != "football")
```

# further edits to data
```{r}

# a duplicate dataset
final_ds3 <- final_ds2
final_ds3$returnerId <- as.numeric(final_ds3$returnerId)
# still has 4185 plays here
fd3p <- final_ds3 %>% dplyr::select(gameId,playId) %>% distinct()


# DELETING PLAYS WITHOUT A RETURNER; removing 766 plays, leaving 3419 plays
final_clean4 <- final_ds3 %>% drop_na(returnerId)
# game and play ids
pnumber <- final_clean4 %>% dplyr::select(gameId,playId) %>% distinct()

# change times to represent number of seconds before half
final_clean4$gameClock <- round(period_to_seconds(final_clean4$gameClock))


# taking out plays without a ball snap label
snap_present <- final_clean4 %>% dplyr::filter(event == "ball_snap") %>% dplyr::select(gameId,playId) %>% distinct() #3397 plays

#taking out plays without a punt label
punt_present <- final_clean4 %>% dplyr::filter(event == "punt") %>% dplyr::select(gameId,playId) %>% distinct() #3415 plays

snap_punt_present <- inner_join(snap_present,punt_present) #3393 plays

#taking out plays without any return event
return_present <- final_clean4 %>% dplyr::filter(event %in% c("punt_received","fair_catch","punt_land")) %>% dplyr::select(gameId,playId) %>% distinct() #3401 plays

 # left with 3375 plays; comprehensive list of game ids and play ids I want to use 
all_present <- inner_join(snap_punt_present,return_present)


```


# Looking at how to analyze a single play; contains the diffs function

```{r}

# EXPERIMENTING WITH A SINGLE PLAY (DEONTE HARRIS TOUCHDOWN)

harris_return <- final_clean4 %>% dplyr::filter(gameId == "2019092211",playId == "218")

str(harris_return)


harris_defs <- harris_return %>% dplyr::filter(kr == "kick",position!= "P",event == "punt_received")

y_order <- c(0,harris_defs$y[order(harris_defs$y)],53.3)

# function that finds gaps between defenders given a frame
diffs <- function(coords){
  
  gaps <- numeric(length(coords)-1)
  
  for(i in 1:length(gaps)){
    gaps[i] <- coords[i+1] - coords[i]
  }
  
  c(median_gap = median(gaps),largest_gap = max(gaps))
}

diffs(y_order)

harris_snap <- harris_return %>% dplyr::filter(event == "ball_snap")


```




# Contains THE GRAND FUNCTION
```{r}

### THE GRAND FUNCTION: Returns crucial variables given a single play

grand_fn <- function(gameId1,playId1,dataset){
  
  play1 <- dataset %>% dplyr::filter(gameId == gameId1,playId == playId1)
  
  # 1.) AT SNAP
  snaptime <- play1 %>% dplyr::filter(event == "ball_snap") %>% dplyr::select(x,y,s,a,dis,o,dir,kr,gunner_status,returner_status,position,team,absoluteYardlineNumber)
  
  # returner info at the snap
  snapreturner <- snaptime %>% dplyr::filter(returner_status == "1")
  

  # obtaining distance from returner to line of scrimmage
      snapdist <- abs(snapreturner$x - snapreturner$absoluteYardlineNumber)

  
    
  # important variables at the snap
    snapvars <- list(responsevar = play1$Response[1],
                gameId = mean(na.omit(play1$gameId)),
                playId = mean(na.omit(play1$playId)),
                scrimdist=as.numeric(snapdist),
                quarter= play1$quarter[1],
                abs_yrdline = mean(na.omit(play1$absoluteYardlineNumber)),
                scorediff = mean(na.omit(play1$scorediff)),
                gmClock = mean(play1$gameClock),
                retid = play1$returnerId[1])
    
  #####################################
  
   # 2.) AT PUNT
punttime <- play1 %>% dplyr::filter(event == "punt") %>% dplyr::select(x,y,s,a,dis,o,dir,kr,gunner_status,returner_status,position,team,absoluteYardlineNumber)
    

    puntvars <- list(klength = mean(na.omit(play1$kickLength)),
                  snap_detail = as.factor(play1$snapDetail[1]),
                  operation_time = play1$operationTime[1],
                  h_time = play1$hangTime[1],
                  ktype = as.factor(play1$kickType[1]),
                  kmatch = as.factor(play1$kickDirectionMatch[1]))
    
    #####################################
    
  # 3.) AT FAIR CATCH / RETURN / PUNT LAND
   
       # frame statistics for all players at time of catch
  returntime <- play1 %>% dplyr::filter(event %in% c("punt_received","fair_catch","punt_land")) %>% dplyr::select(x,y,s,a,dis,o,dir,kr,gunner_status,returner_status,position,team,absoluteYardlineNumber)
    
    # returner statistics at time of catch
    returner_received <- returntime %>% dplyr::filter(returner_status == "1")
    
  # calculating returner's distance traveled from initial spot
    if(nrow(returner_received) > 0 & nrow(snapreturner) > 0){
        ret_dist_trav <- 
          sqrt( (returner_received$x - snapreturner$x)^2 + (returner_received$y - snapreturner$y)^2 )
    } else{
      ret_dist_trav <- NA
    }

      
# frame data for the kick coverage team minus the punter during the punt reception
coverage <- returntime %>% dplyr::filter(kr == "kick",position!="P")


  # a vector of coverage team distances from returner
  current_dist <- numeric(nrow(coverage))
  for(i in 1:nrow(coverage)){
    current_dist[i] <- sqrt( (coverage[i,]$x - returner_received$x)^2 + (coverage[i,]$y - returner_received$y)^2 )
  }

# distance from sideline for returner

 dis_side <- numeric(1)
 if(returner_received$y <= 26.65){
  dis_side <- returner_received$y
 } else{
    dis_side <- 53.3 - returner_received$y
 }

  # average coverage speeds and accelerations
    Med_cspeed = median(coverage$s)
    Med_cacc = median(coverage$a)
  
  # average returner speed and acceleration

    Ret_speed = returner_received$s
    Ret_acc = returner_received$a
    Ret_angle = returner_received$dir

  # ordered vector of y coordinates of kick coverage team to determine gaps
      y_order <- c(0,coverage$y[order(coverage$y)],53.3)
  

  # named vector of new variables
  newvars <- list(Med_cdist = as.numeric(median(current_dist)),
    Closest_cdist=as.numeric(min(current_dist)),
    Med_cspeed = Med_cspeed,
    Med_cacc = Med_cacc,
    Sideline_rdist = dis_side,
    Ret_speed = Ret_speed,
    Ret_acc = Ret_acc,
    Ret_angle = Ret_angle,
    Med_gaps = diffs(y_order)[[1]],
    Max_gaps = diffs(y_order)[[2]])
  
  
                      
      # output vector
   list(c(snapvars,puntvars,newvars,ret_dist_trav = ret_dist_trav))
  

}
### END OF FUNCTION


# FINAL TEST
toy_ds <- data.frame(grand_fn(all_present$gameId[1],all_present$playId[1],final_clean4))


custom <- data.frame(matrix(nrow=nrow(all_present),ncol=26))
names(custom) <- names(toy_ds)

custom$responsevar <- factor(custom$responsevar,levels=c("No","Yes"))
custom$gameId <- as.numeric(custom$gameId)
custom$playId <- as.numeric(custom$playId)
custom$scrimdist <- as.numeric(custom$scrimdist)
custom$quarter <- factor(custom$quarter,levels=c("1","2","3","4","5"))
custom$abs_yrdline <- as.numeric(custom$abs_yrdline)
custom$scorediff <- as.numeric(custom$scorediff)
custom$gmClock <- as.numeric(custom$gmClock)
custom$retid <- as.numeric(custom$retid)
custom$klength <- as.numeric(custom$klength)
custom$snap_detail <- factor(custom$snap_detail,levels=c("<",">","H","L","OK")) 
custom$operation_time <- as.numeric(custom$operation_time)
custom$h_time <- as.numeric(custom$h_time)
custom$ktype <- factor(custom$ktype,levels=c("A","N","R"))
custom$kmatch <- factor(custom$kmatch,levels=c("no","yes"))
custom$Med_cdist <- as.numeric(custom$Med_cdist)
custom$Closest_cdist <- as.numeric(custom$Closest_cdist)
custom$Med_cspeed <- as.numeric(custom$Med_cspeed)
custom$Med_cacc <- as.numeric(custom$Med_cacc)
custom$Sideline_rdist <- as.numeric(custom$Sideline_rdist)
custom$Ret_speed <- as.numeric(custom$Ret_speed)
custom$Ret_acc <- as.numeric(custom$Ret_acc)
custom$Ret_angle <- as.numeric(custom$Ret_angle)
custom$Med_gaps <- as.numeric(custom$Med_gaps)
custom$Max_gaps <- as.numeric(custom$Max_gaps)
custom$ret_dist_trav <- as.numeric(custom$ret_dist_trav)


# initializing dataset; GOAL: consolidate data into one dataset with all predictors for each play, then do modeling

### MAIN DATASET TO BE USED FOR ANALYSIS
  for(i in 1:nrow(all_present)){
    custom[i,] <- data.frame(grand_fn(all_present$gameId[i],all_present$playId[i],final_clean4))
  }



```

# EDA

```{r}

# taking out rows with NAs
custom2 <- custom %>% drop_na()

# making returnerId a factor
custom2$retid <- as.factor(custom2$retid)

# summary of dataset
summary(custom2)

# the response variable
summary(custom2$responsevar)

library(ggplot2)
library(ggcorrplot)

# Visual 1: plot of response variable by quarter
fig1 <- ggplot(aes(x=responsevar,fill=quarter),data=custom2) +
  geom_bar(col="black",position="stack") + xlab("Returner Decision") + ylab("Count") + ggtitle("Return Frequencies for Selected Punt Plays by Quarter, 2018-2020")

fig1

# data to support Visual 1 conclusions
custom2_quarter <- custom2 %>% dplyr::select(responsevar,quarter) %>% group_by(quarter) %>% summarize(count=n())

quarter_freqs <- custom2_quarter$count

custom2_quarter2 <- custom2 %>% dplyr::select(responsevar,quarter) %>% group_by(responsevar,quarter) %>% summarize(count=n()) %>% arrange(quarter)

custom2_quarter2$freq <- numeric(nrow(custom2_quarter2))

for(i in 1:10){
  if(custom2_quarter2$responsevar[i] == "No"){
  custom2_quarter2$freq[i] <- custom2_quarter2$count[i] / 1641
} else{
  custom2_quarter2$freq[i] <- custom2_quarter2$count[i] / 1729
}
}



# Visual 2: this time by kick type; seems like Aussie lead to fewer returns
fig2 <- ggplot(aes(x=responsevar,fill=ktype),data=custom2) +
  geom_bar(col="dark green",position="stack") + xlab("Returner Decision") + ylab("Count") + ggtitle("Return Frequencies for Selected Punt Plays by Quarter, 2018-2020")

fig2

# data supporting Visual 2 conclusions; shows an uneven distribution of punt types across return results
custom2_kt <- custom2 %>% dplyr::select(responsevar,ktype) %>% group_by(responsevar,ktype) %>% summarize(count=n())



# Visual 3: distribution return results based on field position and score differential
fig3 <- ggplot(aes(x=abs_yrdline,y=scorediff,col=responsevar),data=custom2) + 
  geom_point() + 
  ylim(c(-45,45)) + ggtitle("Return Results based on Punt LOS and Score") + xlab("Line of Scrimmage of Punt (from 0-120 yards)") + ylab("Score Differential (Returner's Perspective)")


# Visual 4: checking correlations between numeric predictors
custom2_num <- custom2 %>% dplyr::select(-c(kmatch,ktype,retid,quarter,responsevar,snap_detail))

fig4 <- ggcorrplot(round(cor(custom2_num),2),title="Correlations between Numerical Predictors")


```


# Modeling Process


# Model 1: the full model
```{r}

# MODEL 1: the full model
m1 <- glm(responsevar ~ .-gameId-playId-retid,data=custom2,family="binomial")
summary(m1)


# test train splits for cv; 2/3 train, 1/3 test
library(caret)
set.seed(313)
train_index <- createDataPartition(custom2$responsevar, p = 2/3,list=FALSE)
train <- custom2[train_index, ]
test <- custom2[-train_index, ]

# cross-validation
tr_control <- trainControl(method = "cv", number=5, savePredictions=TRUE)
cv1 <- train(responsevar ~ .-gameId-playId-retid,data=custom2, trControl = tr_control, method = "glm",family = "binomial")

# finding cv accuracy
cv1pred <- predict(cv1, newdata = test)
cv1_table <- table(cv1pred, test$responsevar)
sum(diag(cv1_table)) / sum(cv1_table) # accuracy of 0.838
cv1_table
library(car)
Anova(m1, type = "II", test = "Wald")
# notable variables include scrimdist, klength, h_time, ktype, kmatch, Closest_cdist, Med_cspeed, Med_cacc, Sideline_rdist, Ret_speed, Ret_acc, Med_gaps, Max_gaps, ret_dist_trav

```


# Model 2: reduced model based on correlated predictors and significant predictors
```{r}

# looking at vif factors of individual variables; quarter and gmClock both highest at >4, also they were not significant in the model, so I will remove them
vif(m1)


# MODEL 2: Reduced Version (taking out correlated predictors)

# Decisions based on correlations, variable significance:
# quarter and GmClock to be removed
# ret_dist_trav over Sideline_rdist due to higher variable significance
# Med_gaps over Max_gaps ""
# Med_cspeed over Med_cacc ""

# seeing how correlations change by removing variables
custom3_num <- custom2_num %>% dplyr::select(-c(gameId,playId,gmClock,Sideline_rdist,Max_gaps,Med_cacc))

ggcorrplot(round(cor(custom3_num),2),title="Correlations between Numerical Predictors")


# MODEL 2: the reduced model
m2 <- glm(responsevar ~ .-gameId-playId-retid-quarter-gmClock-Sideline_rdist-Max_gaps,data=custom2,family="binomial")

summary(m2)

# all vifs are below 4
vif(m2)


# cross-validation
cv2 <- train(responsevar ~ .-gameId-playId-retid-quarter-gmClock-Sideline_rdist-Max_gaps,data=custom2, trControl = tr_control, method = "glm",family = "binomial")

# finding cv accuracy
cv2pred <- predict(cv2, newdata = test)
cv2_table <- table(cv2pred, test$responsevar)
sum(diag(cv2_table)) / sum(cv2_table) # accuracy of 0.841496

cv2_table

# anova tables
anova(m2)
Anova(m2, type = "II", test = "Wald")
# notable variables include scrimdist, klength, h_time, ktype, kmatch, Closest_cdist, Med_cspeed, Med_cacc, Ret_speed, Ret_acc, ret_dist_trav

# shows most significant variables in order
important <- Anova(m2, type = "II", test = "Wald") %>% arrange(-Chisq)

# 5 best predictors
rownames(important)[1:5]
```


# Lasso Model

```{r}
library(glmnet)

#convert training data to matrix format
x <- model.matrix(responsevar ~ .-gameId-playId-retid-quarter-gmClock-Sideline_rdist-Max_gaps,data=train)
#convert class to numerical variable
y <- ifelse(train$responsevar=="Yes",1,0)

# finding optimal lambda
cv.lasso <- cv.glmnet(x,y,alpha=1,family="binomial",type.measure = "mse" )
#plot result
plot(cv.lasso)


#best value of lambda
lambda_1se <- cv.lasso$lambda.1se


# test data conversion
x2 <- model.matrix(responsevar ~ .-gameId-playId-retid-quarter-gmClock-Sideline_rdist-Max_gaps,data=test)

# predictions
lasso_pred <- predict(cv.lasso,newx=x2,s=lambda_1se,type="response")

# making predictions back into "Yes" and "No"
lasso_pred2 <- rep("No",nrow(test))
lasso_pred2[lasso_pred > 0.5] <- "Yes"

lasso_tab <- table(pred=lasso_pred2,true=test$responsevar)
sum(diag(lasso_tab)) / sum(lasso_tab) # accuracy of 0.8388

```


# Ridge Model

```{r}
#convert training data to matrix format
xridge <- model.matrix(responsevar ~ .-gameId-playId-retid-quarter-gmClock-Sideline_rdist-Max_gaps,data=train)
#convert class to numerical variable
yridge <- ifelse(train$responsevar=="Yes",1,0)

# finding optimal lambda
cv.ridge <- cv.glmnet(xridge,yridge,alpha=0,family="binomial",type.measure = "mse" )
#plot result
plot(cv.ridge)


#best value of lambda
lambda_1se_ridge <- cv.ridge$lambda.1se

# test data conversion
x2ridge <- model.matrix(responsevar ~ .-gameId-playId-retid-quarter-gmClock-Sideline_rdist-Max_gaps,data=test)

# predictions
ridge_pred <- predict(cv.ridge,newx=x2ridge,s=lambda_1se_ridge,type="response")

# making predictions back into "Yes" and "No"
ridge_pred2 <- rep("No",nrow(test))
ridge_pred2[ridge_pred > 0.5] <- "Yes"

ridge_tab <- table(pred=ridge_pred2,true=test$responsevar)
sum(diag(ridge_tab)) / sum(ridge_tab) # accuracy of 0.833
```

